---
title: "WGBS Spermatogenesis methylation PCA"
author: "Originally Dr. Marius Wöste, edited by Verena Dietrich"
date: "05 31 2023"
updates: "09  29 2023"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
# Rscript -e "rmarkdown::render('PCA_imprinted_genes_Control_Crypto_CZ1_Blood_Sperm.Rmd')"
install.packages("scatterplot3d") # Install
install.packages("rmarkdown")
install.packages("data.table")
install.packages("ggplot2")
install.packages("ggfortify")
install.packages("openxlsx")
library(data.table)
library(stringr)
library(ggplot2)
library(ggfortify)
library(openxlsx)
library(parallel)
library("scatterplot3d")
knitr::opts_chunk$set(echo = TRUE)
```

```{r includes, include=FALSE, echo=FALSE}

source("/data/gitlab/spermatogenesis-methylome/PCA/PCA_analysis/wgbs-methylation/pca.R")
```

## WGBS OAT PCA analysis

Alignment was done with bwa-meth and hg38.
Alignments have been deduplicated using picard tools.
CpG methylation calling has been performed using MethylDackel.

```{r loading-data}

methylationFilePaths <- c(
   "/data/bed-graphs/",
   "/data/bed-graphs/contaminated",
   "/data/bed-graphs/BloodSperm"
)
minCoverage <- 5

methylationValueFiles <- list.files(methylationFilePaths, pattern = ".bedGraph", full.names = TRUE)

methylationValues <- mclapply(methylationValueFiles, function (file) {
  
  values <- fread(file, skip = 1L)

  values <- values[V5 + V6 >= minCoverage]
  
  rownames(values) <- paste(values$V1, values$V2, sep = "-")
  
  return(values)
}, mc.cores = 8)

### Location of imprinted genes

regions_file <- fread(file = "/data/gitlab/spermatogenesis-methylome/PCA/PCA_analysis/Imprinted_genes_location.csv")
regions_imprinted_genes <- regions_file$`Location_(hg38)`
regions_split <- str_split(regions_imprinted_genes, ":|-")
chr <- lapply(regions_split, `[[`, 1)
gene_chr <- str_remove(chr, "chr")
gene_start <- lapply(regions_split, `[[`, 2)
gene_start <- strtoi(gene_start)
gene_end <- lapply(regions_split, `[[`, 3)
gene_end <- strtoi(gene_end)

imprinted_genes <- data.table(V1 = integer(), V2 = integer(), V3 = integer())

for (gene in 1:length(gene_start)){
  position <- as.integer(seq(gene_start[gene],gene_end[gene]))
  chr = rep(strtoi(gene_chr[gene]), length(position))
  imprinted_genes_new <- data.table(V1 = chr, V2 = position, V3 = as.integer(position+1))
  imprinted_genes <- rbindlist(list(imprinted_genes, imprinted_genes_new))
}

imprinted_genes <- unique(imprinted_genes)

### Samples for PCA

sampleNames <- tools::file_path_sans_ext(basename(methylationValueFiles))
sampleNames <- substr(sampleNames, 1, nchar(sampleNames) - nchar("_CpG"))

publicationIDs <- data.table(read.xlsx("/data/gitlab/spermatogenesis-methylome/PCA/PCA_analysis/ENA_PCA_sample_names_plusCrypto.xlsx"))
groupAssociation <- publicationIDs[match(sampleNames, publicationIDs$current_sample_name)]

groupAssociation$Group <- factor(groupAssociation$group_name_for_PCA_1, levels = c("Sperm", "Blood", "Undiff", "Diff", "4C", "1C"), ordered = TRUE)
groupAssociation$Patient <- factor(groupAssociation$patient_name_for_PCA, levels = c("Sperm", "Blood","CTR-1", "CTR-2", "CTR-3", "CZ-1", "CZ-2", "CZ-3"), ordered = TRUE)


names(methylationValues) <- sampleNames
groupAssociation$current_sample_name <- as.character(groupAssociation$current_sample_name)
samplesToKeep <- groupAssociation[!is.na(current_sample_name) & group_name_for_PCA != "not applicable"]$current_sample_name
methylationValues <- methylationValues[samplesToKeep]
groupAssociation <- groupAssociation[current_sample_name %in% samplesToKeep]

### Reduce to imprinted genes

rownames(imprinted_genes) <- paste(imprinted_genes$V1, imprinted_genes$V2, sep = "-")

methylationValues$imprinted_genes <- imprinted_genes

methylationInAllSamples <- Reduce(intersect, lapply(methylationValues, rownames))
methylationValues$imprinted_genes <- NULL 
intersectedValues <- sapply(methylationValues, function (values) values[rownames(values) %in% methylationInAllSamples]$V4 / 100)
```

The following plots are generated for all samples for `r nrow(intersectedValues)` CpG loci where all samples show methylation values. Only loci with minimum coverage of 5 in all samples are considered:

```{r pca, echo=FALSE}


pca <- prcomp(t(intersectedValues))

groupAssociation <- as.data.frame(groupAssociation)

pca1 <- autoplot(pca, data = groupAssociation, fill = "Group", colour = "Group") +

theme_classic() + 
  scale_fill_manual(values = c("grey", "pink", "#001889", "#91008D", "#D24E71",  "#EDA200"))+
    scale_colour_manual(values = c("grey", "pink", "#001889", "#91008D", "#D24E71", "#EDA200")) +
theme(
axis.title.x= element_text(colour = "black", size = 7), 
axis.text.x = element_text(colour = "black", size = 7), 
axis.text.y = element_text(colour = "black", size = 7),
axis.title.y =element_text(colour = "black", size = 7), 
axis.line =   element_line(size = 0.3, colour = "black"), 
panel.background = element_blank()
#legend.position = "none"
)


ggsave("/data/gitlab/spermatogenesis-methylome/PCA/PCA_imprinted_genes_Control_Crypto_CZ1_Blood_Sperm.png", width = 9, height = 4, units = "cm", dpi = 300, pca1)

pca1
```